Refactoring the `nls()` function in R

### 2022-01-19 PkgFromRbase/PkgFromRbase.Rmd
Edits to articles and blog posts

---
title: "Making a package from base R files"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
   - based on contribution from Duncan Murdoch
date: "11/06/2021"
output:
  pdf_document: default
---

@@

This article tries to explain an approach to developing alternative versions of
functions which are in the distributed base of R. Our interest was in developing
improvements to the nls() function and related features in R as part of a Google
Summer of Code project for which Arkajyoti Bhattacharjee is the funded student.
However, nls() has many tentacles involving a number of files and functions that
may or may not be called as `nls()` is executed.

Part of the difficulty in carrying out such development of alternative versions
is that one needs to be able to execute the new variants in parallel with the
existing ones. A heavy-effort approach would be to have separate full sets of
R code and build each system and run them separately. That is, we could have
two versions of R.

Attempts to provide R functions to switch between the sets of working functions
-- e.g., `ReplaceNLS.R` and `RestoreNLS.R` --
were not promising, and not pursued.

Duncan Murdoch suggested, and generously provided an almost-complete prototype of,
a package of interlocked functions that provide the set of `nls()` capabilities.
This package `nlspkg` has been cloned to `nlsalt` which is being modified
to provide the new capabilities or new expression of the code. Moreover, these
capabilities, such as a function `Anyfn()` can be tested side by side by use
of the double colon syntax, namely,

@@

After I downloaded the ZIP file of Duncan's package from his Github repository, 
I built and checked the package. There were two WARNING messages:

@@

- the rest of the "undocumented" functions are C functions (identifiable as
their names have underscore characters in the names). It is these functions
that our "Improvements to nls()" project aims to replace where sensible to
do so with R functions. They may then be documented as needed.

@@

It is worth mentioning that the NAMESPACE file provided by Duncan is an
important part of the packaging. It could be worthwhile to document the
elements in that file needed for a package such as `nlspkg`. In particular,
however, we must note that in renaming the package from `nlspkg` to something
else, the change of name must be mirrored in the line

@@

We also noted that the `S3method()` entries are needed to link the generic
functions to similar modeling functions in R. Thus `anova`, `coef`, etc.
are declared as S3 methods.

@@

to ensure functions were made available to the user. This makes visible all 
the internal C functions, named `C_(something)`. We have suppressed this blanket
declaration and explicitly exported the `nls` and `numericDeric` functions. 
One or two other routines may need to be exported to allow full functioning
of the package.

@@

For version control, the file `.gitignore` has two entries: one is `*.0` and
the other, initially, was `nlspkg.so`. I changed the latter to `*.so` and this
appears to be satisfactory for general use.

# Developing an alternative set of functions and alternative package

The alternative package is being developed by replacing calls to C functions
with R work-alike functions. A preliminary example that already is set up and
has passed preliminary checks is the function `numericDeriv()` from the 
file `R/nls.R` with called elements from the file `src/nls.c`. Note that some
code needs to be commented out or deleted from the C file(s) and headers
to avoid errors when building and checking the alternative.

@@

For Rstudio users, note that each package directory needs its own (name).Rproj
file, created using Rstudio's "File/New Project" etc. dialog. However, the package
directories with their Rproj files can be within the umbrella version control 
project `improvenls` for the Git version control. (We are using a Gitlab
repository.)

@@

While all the above worked well in a Linux environment, when we tried running
the build and CHECK in a Windows environment, we found we got errors. (I normally
run in Linux Mint 20.1, but Arkajyoti Bhattacharjee uses Windows 10. Each of us
has a VirtualBox VM for the alternative OS to allow for checking of our work.)
There is a load error for functions from the BLAS (Basic Linear Algebra Subroutines)
collection in a 32 bit architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
feature. Using the `CMD.exe` terminal window showed that `R CMD build` worked, but
that `R CMD INSTALL` gave the load errors.

@@

```
PKG_LIBS = $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```

I did, however, note that it is not clearly indicated that the file needs to be in 
`src/`, nor (until revealed by `R CMD check`) that the file must be `Makevars.win` and
not `makevars.win`. There is a lot of material on `Makevars` in the Writing R
Extensions manual, but these points do not seem to be presented.

@@

It is a pleasure to note the help and encouragement of Duncan Murdoch, Arkajyoti
Bhattacharjee, Dirk Eddelbuettel and Brett Klamer.

>>>>

---
title: "Making a package from base R files"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
   - based on communications from Duncan Murdoch
date: "11/06/2021, revised 19/01/2022"
bibliography: ../BibSupport/ImproveNLS.bib
output:
  pdf_document: default
---

@@

This article tries to explain an approach to developing alternative versions of
functions which are in the distributed base of R. Our interest was in developing
improvements to the nls() function and related features in R as part of a Google
Summer of Code project for which Arkajyoti Bhattacharjee was the funded student.
However, nls() has many tentacles involving a number of files and functions that
may or may not be called as `nls()` is executed.

Part of the difficulty in carrying out such development of alternative versions
is that one needs to be able to execute the new variants in parallel with the
existing ones. A heavy-effort approach would be to have separate full sets of
R code and build each system and run them separately. That is, we want to have
two or more versions of R in the same computing system.

There are several ways to do this:

- build or install the two versions of R in separate directory trees and
  invoke them with separate launch scripts or programs, such as described
  in https://support.rstudio.com/hc/en-us/articles/215488098-Installing-multiple-versions-of-R-on-Linux ;

- use virtual machines with different R versions;

- run the different R versions in containers, e.g., @Rocker2017;

- move the required functionality to a package that can be loaded as needed
  to mask the base functionality and replace it with the functions we wish
  to test or use. This last approach is discussed here.

We did attempt to provide R functions to switch between alternative sets of working functions
-- e.g., `ReplaceNLS.R` and `RestoreNLS.R` -- but this approach 
was not promising, and not pursued.

Duncan Murdoch suggested, and generously provided an almost-complete prototype of,
a package of interlocked functions that provide the set of `nls()` capabilities.
This package `nlspkg` was copied to `nlsalt` then modified
to provide the new capabilities or new expression of the code. The new 
capabilities or code, such as a function `Anyfn()` can be tested side by side with 
a "standard" version (if one exists) by use of the double colon syntax, namely,

@@

After we downloaded the ZIP file of Duncan's package from his Github repository, 
we built and checked the package. There were two WARNING messages:

@@

- the rest of the "undocumented" functions are C functions (identifiable easily
as their names contain underscore characters). It is these functions
that our "Improvements to nls()" project aims to replace where it is sensible to
do so with R functions. They may then be documented as needed.

@@

It is worth mentioning that the NAMESPACE file provided by Duncan is an
important part of the packaging. It could be worthwhile to document each
elements in that file needed for a package such as `nlspkg`, but at the
time of writing, this is yet to be carried out. However, we must note 
that in renaming the package from `nlspkg` to something
else, the change of name must be mirrored in the line

@@

We also noted that the `S3method()` entries of the NAMESPACE files 
are needed to link the generic
functions to similar modeling functions in R. Thus `anova`, `coef`, etc.
are declared as S3 methods.

@@

to ensure functions were made available to the user. This makes visible all 
the internal C functions, named `C_(something)`. We have suppressed this blanket
declaration and explicitly exported the `nls` and `numericDeriv` functions. 
One or two other routines may need to be exported to allow full functioning
of the package.

@@

For version control, the file `.gitignore` has two entries: one is `*.0` and
the other, initially, was `nlspkg.so`. We changed the latter to `*.so` and this
appears to be satisfactory for general use.

# Developing an alternative set of functions and an alternative package

We wished to build our alternative package by replacing calls to C functions
with R work-alike functions. A preliminary example that already is set up and
has passed preliminary checks is the function `numericDeriv()` from the 
file `R/nls.R` with called elements from the file `src/nls.c`. Note that some
code needs to be commented out or deleted from the C file(s) and headers
to avoid errors when building and checking the alternative package.

@@

For Rstudio users, note that each package directory needs its own (name).Rproj
file, created using Rstudio's "File/New Project" etc. dialog. However, the package
directories with their Rproj files can be within the umbrella version control 
project `improvenls` for the Git version control. (We used a Gitlab
repository.)

@@

While all the above worked well in a Linux environment, when we tried running
the build and CHECK in a Windows environment, we found errors. (JN normally
runs in Linux Mint 20.x, but AB uses Windows 10. Each of us
has a VirtualBox VM for the alternative OS to allow for checking of our work.)
There is a load error for functions from the BLAS (Basic Linear Algebra Subroutines)
collection in a 32 bit Windows architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
feature. Using the `CMD.exe` terminal window showed that `R CMD build` worked, but
that `R CMD INSTALL` gave the load errors.

@@

```
PKG_LIBS = $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```
We did, however, note that it is not clearly indicated that this file needs to be in 
`src/`, nor (until revealed by `R CMD check`) that the file must be `Makevars.win` and
not `makevars.win`. There is a lot of material on `Makevars` in the Writing R
Extensions manual (@R-extns), but these points do not seem to be presented, or at least are
not obvious.

Our packaged version of nls() from July 2021 is at 
https://gitlab.com/nashjc/improvenls/-/tree/master/nlspkg 
Note that this is drawn from the version of R of that date (R 4.1.0), so will not necessarily reflect
the distributed version of R when this document is read.

@@

It is a pleasure to note the help and encouragement of Duncan Murdoch, 
Dirk Eddelbuettel and Brett Klamer.

# References

### 2022-01-19 RefactoringNLS/RefactoringNLS.Rmd
Edits to articles and blog posts

The project output is available in the Git repository https://gitlab.com/nashjc/improvenls
?? Arkajyoti -- do you want to change to Github

@@

These are documents used to discuss particular aspects of our work. These are part
of the repository https://gitlab.com/nashjc/improvenls.

@@

- ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
  project, possibly with wider application to nonlinear least squares in general

- MachineSummary.Rmd: an informal investigation of ways to report the characteristics
  and identity of machines running tests. `MachID.R` offers a concise summary function
  to identify a particular computational system used for tests. 
  Note that there remains an open issue that the get_ram() for Windows 10 in the `benchmarkme`
  package may report the RAM size as "NA". We are collaborating with Colin Gillespie, 
  maintainer of `benchmarkme` to seek a workaround.

@@

Hans Werner Borchers was helpful in developing the proposal for this project.

>>>>

The project output is available in the Git repository https://github.com/nashjc/RNonlinearLS .

@@

These are documents used to discuss particular aspects of our work. These are part
of the repository https://github.com/nashjc/RNonlinearLS.

@@

- ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
  project, possibly with wider application to nonlinear least squares in general.

- MachineSummary.Rmd: an investigation of ways to report the characteristics
  and identity of machines running tests. `MachID.R` offers a suggested concise summary function
  to identify a particular computational system used for tests.

@@

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work.
